Tidy Data

### 2014-01-10 tidy-data.tex
Changes from Garrett

The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (\code{person}, \code{treatment}, \code{result}) and six observations (two for each person). It has two values of the \code{treatment} variable (a and b), three values of the \code{person} variable, and either five or six values of the \code{result} variable, depending on how you think about the missing value.  We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Table~\ref{tbl:preg-tidy}.

@@

Tidy data makes it easy for an analyst or a computer to extract needed variable because it provides a standard way of structuring a dataset. Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies for different variables. This slows analysis and invites errors. If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a standard, simple way. Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.

>>>>

The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} contains three variables that comprise 10 values. The data contains a \code{person} variable with three values (John, Mary, and Jane), a \code{treatment} variable with two values (a and b), and a \code{result} variable with has five or six values depending on how you think of the missing value (-, 16, 3, 2, 11, 1). The layout of Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} makes the number of values clear, but obscures their meaning.

The meaning of the dataset relies on how these values are related to each other. The \code{result} variable was collected over the course of six observations: each \code{person} was observed twice, once for each \code{treatment}. These observations give meaning to the specific values of result. For example, the value 16 is meaningful in the context of the observation: \emph{Jane} scored \emph{16} when she took treatment \code{a}. Table~\ref{tbl:preg-tidy} organizes the dataset around the six observations, making the meaning of the individual values more clear.

@@

Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset. Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables. This slows analysis and invites errors. If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a standard, simple way. Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.
